---
title: "R Notebook"
output: html_notebook
---

Libraries
```{r}
library(readxl)
library(readr)
library(MatchIt)
library(dplyr)
library(broom)
library(car)
library(pROC)
library(ResourceSelection)
library(foreign)
library(tidyverse)
library(knitr)
library(ggplot2)
```

Import data
```{r}
Participant_table_with_related_disorders <- read_excel("../../Data/Participant_table with related disorders.xlsx")
PSP_w_exposures <- read_csv("../../Data/PSP_w_exposures.csv")
PSP_assessment_centre <- read_csv("../../Data/PSP_assessment_centre.csv")
Control_w_exposures <- read_csv("../../Data/Control_w_exposures.csv")
Controls_age_and_sex <- read_csv("../../Data/Controls_age_and_sex.csv")
PMH_and_meds <- read_csv("../../Data/Med history and meds self reported.csv")
Metabolic_syndrome <- read_csv("../../Data/metabolic_syndrome_participant.csv")
```

Combine groups into PSP and healthy cohorts, then combine into single dataframe

```{r}
Controls_all <- merge(Control_w_exposures,
                      Controls_age_and_sex,
                      by = "Participant ID")
Controls_all <- merge(Controls_all,
                      PMH_and_meds,
                      by = "Participant ID")
PSP_all <- merge(PSP_w_exposures,
                 Participant_table_with_related_disorders,
                 by = "Participant ID")
PSP_all <- merge(PSP_all,
                 PSP_assessment_centre,
                 by = "Participant ID")
PSP_all_w_pmh <- merge(PSP_all,
                 PMH_and_meds,
                 by = "Participant ID")
if ("UK Biobank assessment centre | Instance 0.y" %in% names(PSP_all)) { #Fix name discrepancy
  PSP_all <- PSP_all_w_pmh %>%
    rename(`UK Biobank assessment centre | Instance 0` =
             `UK Biobank assessment centre | Instance 0.y`)
}

psp  <- PSP_all %>% mutate(case = 1L) %>%
  filter(`Years from recruitment to first report of PSP` >= 0)
ctrl <- Controls_all %>% mutate(case = 0L)
df <- bind_rows(psp,ctrl)
df <- merge(df, Metabolic_syndrome, by = "Participant ID")
```

Make matched control group, matched on age, sex, and assessment center, at 10:1 ratio
```{r}
m <- matchit(
  case ~ `Sex` + `Age at recruitment` + `UK Biobank assessment centre | Instance 0`,
  data = df,
  method = "nearest",
  distance = "glm",
  ratio = 10
)

summary(m)

matched_all <- match.data(m)
```


Analysis - see if CVD is more common in PSP than controls (by self report)
```{r}
cvd_table <- matched_all %>%
  select(`Participant ID`, 
         `case`,
         cvd_var = `Vascular/heart problems diagnosed by doctor | Instance 0`) %>%
  filter(cvd_var != "Prefer not to answer") %>%
  mutate(has_cvd = if_else(cvd_var == "None of the above", 0L, 1L))

cvd_summary <- cvd_table %>%
  group_by(case) %>%
  summarise(
    n = n(),
    n_with_cvd = sum(has_cvd),
    percent_with_cvd = 100 * mean(has_cvd),
    .groups = "drop"
  )

cvd_tab <- table(cvd_table$case, cvd_table$has_cvd)
cvd_tab
chisq.test(cvd_tab)

```
Analysis - see if self-reported HTN is more common in PSP than controls (run the CVD one above first)
```{r}
htn_table <- cvd_table %>%
  mutate(has_htn = if_else(
    str_detect(cvd_var, regex("High blood pressure", ignore_case = TRUE)), 1L, 0L))

htn_summary <- htn_table %>%
  group_by(case) %>%
  summarise(
    n = n(),
    n_with_htn = sum(has_htn),
    percent_with_htn = 100 * mean(has_htn),
    .groups = "drop"
  )

htn_tab <- table(htn_table$case, htn_table$has_htn)
htn_tab
chisq.test(htn_tab)

```
Analysis - see if self-reported diabetes is more common in PSP than controls
```{r}
dm_table <- matched_all %>%
  select(`Participant ID`, 
         `case`,
         dm_var = `Diabetes diagnosed by doctor | Instance 0`) %>%
  filter(dm_var != "Prefer not to answer") %>%
  filter(dm_var != "Do not know") %>%
  mutate(has_dm = if_else(dm_var == "No", 0L, 1L))

dm_summary <- dm_table %>%
  group_by(case) %>%
  summarise(
    n = n(),
    n_with_dm = sum(has_dm),
    percent_with_dm = 100 * mean(has_dm),
    .groups = "drop"
  )

dm_tab <- table(dm_table$case, dm_table$has_dm)
dm_tab
chisq.test(dm_tab)

```
Analysis - see if metabolic syndrome markers are different between PSP and controls
```{r}
metsyn_table <- matched_all %>%
  select(
    ID = `Participant ID`, 
    case = `case`,
    BMI = `Body mass index (BMI) | Instance 0`,
    LDL = `LDL direct | Instance 0`,
    LpA = `Lipoprotein A | Instance 0`,
    A1c = `Glycated haemoglobin (HbA1c) | Instance 0`,
    htn_date = `Date I10 first reported (essential (primary) hypertension)`,
    iddm_date = `Date E10 first reported (insulin-dependent diabetes mellitus)`,
    niddm_date = `Date E11 first reported (non-insulin-dependent diabetes mellitus)`,
    osdm_date = `Date E13 first reported (other specified diabetes mellitus)`,
    udm_date = `Date E14 first reported (unspecified diabetes mellitus)`
  ) %>%
  mutate(earliest_dm_date = pmin(iddm_date, niddm_date, osdm_date, udm_date, na.rm=TRUE)) %>%
  mutate(any_htn = ifelse(is.na(htn_date), 0, 1)) %>%
  mutate(any_dm = ifelse(is.na(earliest_dm_date), 0, 1))

#Display descriptive data
desc_table <- metsyn_table %>%
  group_by(case) %>%
  summarise(
    n = n(),
    
    # BMI
    BMI_mean   = mean(BMI, na.rm = TRUE),
    BMI_sd     = sd(BMI, na.rm = TRUE),
    BMI_median = median(BMI, na.rm = TRUE),
    BMI_IQR    = IQR(BMI, na.rm = TRUE),
    BMI_missing = sum(is.na(BMI)),
    
    # LDL
    LDL_mean   = mean(LDL, na.rm = TRUE),
    LDL_sd     = sd(LDL, na.rm = TRUE),
    LDL_median = median(LDL, na.rm = TRUE),
    LDL_IQR    = IQR(LDL, na.rm = TRUE),
    LDL_missing = sum(is.na(LDL)),
    
    # LpA
    LpA_mean   = mean(LpA, na.rm = TRUE),
    LpA_sd     = sd(LpA, na.rm = TRUE),
    LpA_median = median(LpA, na.rm = TRUE),
    LpA_IQR    = IQR(LpA, na.rm = TRUE),
    LpA_missing = sum(is.na(LpA)),
    
    # A1c
    A1c_mean   = mean(A1c, na.rm = TRUE),
    A1c_sd     = sd(A1c, na.rm = TRUE),
    A1c_median = median(A1c, na.rm = TRUE),
    A1c_IQR    = IQR(A1c, na.rm = TRUE),
    A1c_missing = sum(is.na(A1c)),
    
    # Binary vars: percentage
    any_htn_percent = 100 * mean(any_htn == 1, na.rm = TRUE),
    any_dm_percent  = 100 * mean(any_dm == 1, na.rm = TRUE)
  )

print(desc_table)

#Univariate analysis

results_list <- list()

# Continuous variables ----------------
numeric_vars <- c("BMI", "LDL", "LpA", "A1c")

for (var in numeric_vars) {
  formula <- as.formula(paste(var, "~ case"))
  
  # Wilcoxon test (robust to non-normality)
  test <- wilcox.test(formula, data = metsyn_table)
  
  results_list[[var]] <- tibble(
    variable = var,
    test_type = "Wilcoxon rank-sum",
    statistic = unname(test$statistic),
    p_value = test$p.value
  )
}

# Binary variables --------------------
binary_vars <- c("any_htn", "any_dm")

for (var in binary_vars) {
  tab <- table(metsyn_table$case, metsyn_table[[var]])
  
  if (any(tab < 5)) {
    test <- fisher.test(tab)
    method <- "Fisher's exact"
  } else {
    test <- chisq.test(tab)
    method <- "Chi-square"
  }
  
  results_list[[var]] <- tibble(
    variable = var,
    test_type = method,
    statistic = ifelse(method == "Chi-square", unname(test$statistic), NA),
    p_value = test$p.value
  )
}

# Combine results
univariate_results <- bind_rows(results_list)

print(univariate_results)

#Plot numerical results
for (v in numeric_vars) {
  p <- ggplot(metsyn_table, aes(x = factor(case), y = .data[[v]], fill = factor(case))) +
    geom_violin(trim = FALSE, alpha = 0.6) +
    geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.7) +
    labs(x = "Case status", y = v,
         title = paste("Distribution of", v, "by case"),
         fill = "Case") +
    theme_minimal(base_size = 14)
  
  print(p)
}
```

Analysis - number of diagnoses and meds
```{r}
numdx_table <- matched_all %>%
  select(`Participant ID`, 
         `case`,
         numdx_var = `Number of self-reported non-cancer illnesses | Instance 0`,
          nummeds_var = `Number of treatments/medications taken | Instance 0`)

numdx_table %>%
  group_by(case) %>%
  summarise(
    n = n(),
    median_numdx = median(numdx_var, na.rm = TRUE),
    IQR_numdx = IQR(numdx_var, na.rm = TRUE),
    mean_numdx = mean(numdx_var, na.rm = TRUE),  # just for context
    sd_numdx = sd(numdx_var, na.rm = TRUE)
  )

numdx_table %>%
  group_by(case) %>%
  summarise(
    n = n(),
    median_nummeds = median(nummeds_var, na.rm = TRUE),
    IQR_nummeds = IQR(nummeds_var, na.rm = TRUE),
    mean_nummeds = mean(nummeds_var, na.rm = TRUE),  # just for context
    sd_nummeds = sd(nummeds_var, na.rm = TRUE)
  )

wilcox.test(numdx_var ~ case, data = numdx_table)
wilcox.test(nummeds_var ~ case, data = numdx_table)
```

Analysis - education level (12 years as cutoff for higher education)
```{r}
edu_table <- matched_all %>%
  select(`Participant ID`, 
         `case`,
          edu_var = `Qualifications | Instance 0`) %>%
          mutate(
            high_edu = case_when(
              str_detect(edu_var, "College or University degree") ~ 1L,
              str_detect(edu_var, "A levels/AS levels or equivalent") ~ 1L,
              str_detect(edu_var, "NVQ or HND or HNC or equivalent") ~ 1L,
              str_detect(edu_var, "Other professional qualifications") ~ 1L,
              str_detect(edu_var, "None of the above") ~ 0L,
              str_detect(edu_var, "Prefer not to answer") ~ NA_integer_,
              TRUE ~ 0L
            )
          ) %>%
    filter(!is.na(high_edu))

# Summarize counts and percentages
edu_summary <- edu_table %>%
  group_by(case) %>%
  summarise(
    n = n(),
    n_high = sum(high_edu == 1),
    percent_high = 100 * mean(high_edu == 1),
    .groups = "drop"
  )

print(edu_summary)

# 2x2 contingency table
edu_tab <- table(edu_table$case, edu_table$high_edu)
edu_tab

# Chi-square test (works if expected counts >= 5)
chisq.test(edu_tab)
```

Analysis - smoking (based on pack-years only)
```{r}
smoking_table <- matched_all %>%
  select(`Participant ID`, 
         `case`,
         packyrs_var = `Pack years of smoking | Instance 0`) %>%
  mutate(packyrs_var = replace_na(packyrs_var, 0)) %>%
  mutate(never_smoker = if_else(packyrs_var == 0, 1L, 0L))

smoking_table %>%
  group_by(case) %>%
  summarise(
    n = n(),
    median_packyrs = median(packyrs_var, na.rm = TRUE),
    IQR_packyrs = IQR(packyrs_var, na.rm = TRUE),
    mean_packyrs = mean(packyrs_var, na.rm = TRUE),  # just for context
    sd_packyrs = sd(packyrs_var, na.rm = TRUE)
  )

# Summarize percentages by case status
smoking_summary <- smoking_table %>%
  group_by(case) %>%
  summarise(
    n = n(),
    n_never = sum(never_smoker),
    percent_never = 100 * mean(never_smoker),
    .groups = "drop"
  )

print(smoking_summary)

# 2x2 contingency table
smoking_tab <- table(smoking_table$case, smoking_table$never_smoker)
smoking_tab

# Statistical test
wilcox.test(packyrs_var ~ case, data = smoking_table)
chisq.test(smoking_tab)
```
Analysis - air pollution
```{r}
pollution_table <- matched_all %>%
  select(`Participant ID`, 
         `case`,
         pollution_var = `Particulate matter air pollution (pm10); 2007`)

pollution_table %>%
  group_by(case) %>%
  summarise(
    n = n(),
    median_pollution = median(pollution_var, na.rm = TRUE),
    IQR_pollution = IQR(pollution_var, na.rm = TRUE),
    mean_pollution = mean(pollution_var, na.rm = TRUE),  # just for context
    sd_pollution = sd(pollution_var, na.rm = TRUE)
  )


wilcox.test(pollution_var ~ case, data = pollution_table)
t.test(pollution_var ~ case, data = pollution_table)
```

Import GP records
```{r}
CTV_library <- read.dbf("C:/Users/Charlie Zhao/OneDrive - Mass General Brigham/Research/UK Biobank/nhs_readbrowser_25.0.0_20180401000001/Standard/V3/Term.dbf")
CTV_tempance <- read.dbf("C:/Users/Charlie Zhao/OneDrive - Mass General Brigham/Research/UK Biobank/nhs_readbrowser_25.0.0_20180401000001/Standard/V3/TEMPANCE.dbf")
CTV_ancestor <- read.dbf("C:/Users/Charlie Zhao/OneDrive - Mass General Brigham/Research/UK Biobank/nhs_readbrowser_25.0.0_20180401000001/Standard/V3/ANCESTOR.dbf")
CTV_hier <- read.dbf("C:/Users/Charlie Zhao/OneDrive - Mass General Brigham/Research/UK Biobank/nhs_readbrowser_25.0.0_20180401000001/Standard/V3/HIER.dbf")
CTV_temphier <- read.dbf("C:/Users/Charlie Zhao/OneDrive - Mass General Brigham/Research/UK Biobank/nhs_readbrowser_25.0.0_20180401000001/Standard/V3/TEMPHIER.dbf")
CTV_concept <- read.dbf("C:/Users/Charlie Zhao/OneDrive - Mass General Brigham/Research/UK Biobank/nhs_readbrowser_25.0.0_20180401000001/Standard/V3/CONCEPT.dbf")
CTV_term198 <- read.dbf("C:/Users/Charlie Zhao/OneDrive - Mass General Brigham/Research/UK Biobank/nhs_readbrowser_25.0.0_20180401000001/Standard/V3/TERM198.dbf")
CTV_ICD10 <- read.dbf("C:/Users/Charlie Zhao/OneDrive - Mass General Brigham/Research/UK Biobank/nhs_readbrowser_25.0.0_20180401000001/Standard/V3/ICD10.dbf")
CTV_Compkey <- read.dbf("C:/Users/Charlie Zhao/OneDrive - Mass General Brigham/Research/UK Biobank/nhs_readbrowser_25.0.0_20180401000001/Standard/V3/COMPKEY4.dbf")


PSP_GPlevel <- read_csv("C:/Users/Charlie Zhao/OneDrive - Mass General Brigham/Research/UK Biobank/Data/PSP-w-GP-only_GPlevel.csv")

```

Join
```{r}
PSP_GPlevel <- PSP_GPlevel %>%
  left_join(
    CTV_library %>% select(TERMID, TERM30),
    by = c("CTV3 (Read v3) code" = "TERMID")
  ) %>%
  rename(Term = TERM30)
```


